---
title: "Lesson 2 Visualization"
author: "Jolynn Schmidt"
date: "June 13, 2018"
output: html_document
---

In this exercise we will start by looking at the characteristics of our dataset. All definitions and data come from the original Udacity lesson and can be found [here.](https://www.udacity.com/api/nodes/1480558581/supplemental_media/lesson2pdf/download "Lesson2.pdf")


###Data
We start by assigning our data to Petals, a list of petal counts from flowers.
```{r}
Petals <- c(15, 16, 17,
            16, 21, 22,
            15, 16, 15,
            17, 16, 22,
            14, 13, 14,
            14, 15, 15,
            14, 15, 16,
            10, 19, 15,
            15, 22, 24,
            25, 15, 16)
```


###Frequency
The frequency of a data set is the number of times a certain outcome occurs.

```{r}
# Find the most frequent petal count.
# to do this we create a table from Petals, sort it in decreasing order, and get the name of the first item.

MostFreq <- names(sort(table(Petals),decreasing=TRUE)[1])
print(paste0("The most frequent petal count is: ", MostFreq))
```

```{r}
# Find the frequency of flowers with 15 petals.
# We can use the same method as above but without the names
PetalFreq <- sort(table(Petals),decreasing=TRUE)[1]
print(paste0("The frequency of flowers with 15 petals is: ", PetalFreq))
```


###Proportions
A proportion is the fraction of counts over the total sample.

```{r}
# Find the proportion of flowers with 15 petals

TotalSample <- length(Petals)
PetalProp <- PetalFreq/TotalSample
print(paste0("The proportion of flowers with 15 petals is: ", PetalProp))
```


###Percentage
A proportion can be turned into a percentage by multiplying the proportion by 100.

```{r}
# Find the percentage of flowers with 15 petals

PetalPerc <- PetalProp * 100
print(paste0("The percentage of flowers with 15 petals is: ", PetalPerc, "%"))
```


###Histogram
A histogram is a graphical representation of the distribution of data, discrete intervals (bins) are decided upon to form widths for our boxes. R has the hist() function to make basic histograms. Here are some very simple examples.

```{r}
# Create a histogram with a bin size of 2
hist(Petals, breaks = (length(Petals)/2), col = "blue")
```

```{r}
# Create a histogram with a bin size of 5
bins <- c(10,15,20,25,30)
#hist(Petals, breaks = (length(Petals)/5))
hist(Petals, breaks = bins, col = "orange")
```


###Skew
**Positive Skew** - A positive skew is when outliers are present along the right most end of the distribution. The histogram for Petals is an example of a positive skew

**Negative Skew** - A negative skew is when outliers are present along the left most end of the distribution
